---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r loadingdata, results="hide", message=FALSE}
#A. Set working ddirectory
setwd("~/Data Scientist/Projects/RepRes")
#B. clear memory if required
#rm(list=ls())

#C. set url and file name
furl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
fname <- "repdata_data_activity.zip"
fpath <-  file.path("activity.csv")

#D. downlad the file
if (!file.exists(fname)) {
    download.file(furl, destfile = fname);
}

#E. unzip if not already unzipped
if (!file.exists(fpath))   unzip(fname,exdir='.')  else {message(paste(fpath, 'already exists'))}

#F.Store data into variable
alldata <- read.csv(fpath,stringsAsFactors = FALSE)

#G. convert date column into Date
alldata$date <- as.Date(alldata$date)

```


## What is mean total number of steps taken per day?

Generate the graphing data using dplyr library grouping on date
```{r calculatetotstepsmeanmedian, results="hide", message=FALSE}
#load library 
library(dplyr)


#generate graphing data
graphdata <- alldata %>% 
    group_by(date) %>% 
        summarize(total.steps.per.day=sum(steps,na.rm=TRUE))
```
Generate the histogram of the total steps per day and calculate mean and median
```{r outputhistogramandtable}
#Histogram of the total steps
hist(graphdata$total.steps.per.day, col='blue',main='Histogram of total steps taken each day', xlab='Total steps per day', ylab='Frequency')

#Find mean and medain
smean <- mean(graphdata$total.steps.per.day, na.rm=TRUE)
smedian <- median(graphdata$total.steps.per.day ,na.rm=TRUE)
smean <- format(round(smean,2), nsmall=2)
smedian <- format(round(smedian,2), nsmall=0)


```
The mean of total number of steps per day is `r smean`
The median of total number of steps per day is `r smedian`

## What is the average daily activity pattern?
Generate the graphing data grouping over interval using dplyr
```{r graphdatalineplot, message=FALSE}
#generate graphing data
graphdata <- alldata %>% 
    group_by(interval) %>% 
        summarize(avg.steps.per.interval=mean(steps,na.rm=TRUE))
```
Plot the line plot on the average steps per interval
```{r lineplot}
#Generate the line plot
with(graphdata, plot(x=interval, y=avg.steps.per.interval, type="l",col="blue", main="Average steps per interval across all days", xlab="5 Minute interval index", ylab="Average steps per interval / day "))
```

Find the interval with maximum average

```{r findmaxinterval}
maxsteps <- max(graphdata$avg.steps.per.interval, na.rm=TRUE)
maxint <- graphdata[graphdata$avg.steps.per.interval==maxsteps,]$interval
```
The interval with the maximum average steps is `r maxint`

## Imputing missing values
Generate median values for each interval across all days
```{r generatemedstepsforimpute}
graphdata <- alldata %>% 
    group_by(interval) %>% 
        summarize(medsteps=median(steps,na.rm=TRUE))
```

merge median value to all data per interval
```{r mergemedianwithalldata}
alldata <- merge(x=alldata, y=graphdata, by="interval")
```
Find the column with NA
```{r findNA}
t <- as.data.frame(lapply(alldata, is.na))
sapply(t,sum)
```

Impute na values in steps as it has NA's
```{r createnewimputeddata}
newdata <- mutate(alldata,steps=ifelse(is.na(steps),medsteps,steps))

```

Calculating histogram, mean and median of total steps each day
```{r imputedtable}
#generate graphing data
graphdata <- newdata %>% 
    group_by(date) %>% 
        summarize(total.steps.per.day=sum(steps,na.rm=TRUE))

#Histogram of the total steps
hist(graphdata$total.steps.per.day, col='blue',main='Histogram of total steps taken each day with imputed data', xlab='Total steps per day', ylab='Frequency')

#Find mean and medain
smean <- mean(graphdata$total.steps.per.day, na.rm=TRUE)
smedian <- median(graphdata$total.steps.per.day ,na.rm=TRUE)
smean <- format(round(smean,2), nsmall=2)
smedian <- format(round(smedian,2), nsmall=0)
```
The new mean of total number of steps per day is `r smean` and has increased over the non imputed data.
The new median of total number of steps per day is `r smedian` and remains the same as the non imputed data primarily because the median was used to impute.


## Are there differences in activity patterns between weekdays and weekends?
Create vector with weekday or weekend
```{r findweekday}
wd <- ifelse(weekdays(newdata$date, abbreviate=TRUE) %in% c("Sat", "Sun"), "weekend", "weekday")
```

Convert to factor and add to the newdata
```{r addfactorweekday}
wd <- factor(wd)
newdata <- cbind(newdata,wd)
```

Generate graph data
```{r createpanelgraphdata}
graphdata <- newdata %>% group_by(wd,interval) %>% summarize(avgsteps=mean(steps))

```
Create plots in panels using ggplot2
```{r createplot}
library(ggplot2)
g <- ggplot(graphdata, aes(interval, avgsteps)) +geom_line(color="blue")
g <- g + facet_grid(wd~.)
g <- g + labs(title="Comparison of Weekend and Weekday average interval steps", x="Interval", y="Average number of steps")
g <- g + theme(strip.text = element_text(size = 12, colour = "black", angle = 90))
g
```

Create the .md and .html file using knit2html
```{r}
#knit2html("PA1_template.Rmd", force_v1 = TRUE)
```

